Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-25 3-Introduction.tex

TODO
MDE has traditionally adopted a two-level,top-down meta-modelling approach, where meta-models reside in a certain meta-level, and models are created one meta-level below by using types from the metamodel (cite MOF and EMF).

This strict approach shows its limitations when modeling complex domains which require more than one level of specialization, e.g., to adapt the model to application sub domains. Indeed, the two-level approach fails to acknowledge the existence of two different forms of classification (ontological vs linguistic) and the duality class-object, leading to the introduction of accidental complexity when dealing with the aforementioned complex domains.

In order to tackle this problem multi-level modeling have been introduced. Contrary to the strict approach, multi-level modeling advocates the use of a flexible number of levels as well as a flexiblzation of the instantiation concept.

This approach is gaining interest lately within the modeling community as shown by the MULTI level-modeling workshop and the daghtul seminar and number of tools. Taking advantage of this vibrant state of affairs, and in order to foster discussion and enable comparison between competing approaches, a multi-level modeling challenge has been created. This paper is a response to the latest multi-level modeling challenge which consist of modeling processes. We do so by using model federation and the OpenFlexo framework.

Model federation is a modeling approach in which...

>>>>

Model-driven engineering (MDE) has traditionally adopted a strict hierarchical two-level approach, where metamodels reside in a certain meta-level, and models are created one meta-level below by using types from the metamodel. Notable examples of this approach are the widespread Eclipse Modeling Framework (EMF)~\citep{emf} and the Object Management Group Meta-Object Facility(MOF)~\citep{omg2013mof}.

This strict approach shows its limitations when modeling complex domains requiring more than one level of specialization, e.g., to adapt the model to application sub domains. Indeed, the two-level approach fails to acknowledge and support: 1) the existence of different forms of classification and/or instantiation (e.g., ontological vs linguistic); and 2) the duality type-object for model elements. Therefore, while it may still be used for the aforementioned complex modeling tasks, the strict approach entails the introduction of accidental complexity in both the modeling process and the resulting modeling artefacts.

In order to tackle this problem, the \emph{multi-level} modeling paradigm have been introduced. Contrary to the strict approach, multi-level modeling advocates the use of a flexible number of levels as well as a flexibilization of the relations between levels.

This approach is gaining interest lately within the modeling community as shown by the MULTI level-modeling workshop and the daghtul seminar~\citep{almeida2018multi} and number of tools. Taking advantage of this vibrant state of affairs, and in order to foster discussion and enable comparison between competing approaches, a multi-level modeling challenge has been created. This paper is a response to the latest multi-level modeling challenge which consist of modeling processes. We do so by using model federation and the OpenFlexo framework.

Model federation is a modeling approach in which loosely couple reified models. Links among models can be within a same level of abstraction or across levels of abstractions which gives, as our answer to the challenge shows, a great flexibility.
